Stochastic Approximation Monte Carlo with a Dynamic Update Factor (SAD)

### 2019-01-23 papers/histogram/sad.tex

Wang and Landau introduced an algorithm that uses an update factor and
a statistical histogram to compute the density of states of a given
system~\cite{wang2001determining, wang2001efficient}.  While the method
is incredibly powerful, it has several user-defined parameters.  This
can make its application to a variety of systems something of an
art-form since the user needs to determine the ideal parameters for the
particular system being studied.  Also, detailed balance is violated
(although only for short periods of time), ensuring that convergence is
not guaranteed.  This raises several questions such as when the method
does converge, what rate will the method converge to the correct
density of states. Also, how can the user appropriately decide what
parameters to choose so that the algorithm solves the problem in the
most ideal way.

%Transition Matrix Monte Carlo~\cite{wang1999transition,
%swendsen1999transition, fitzgerald2000monte} became an attractive
%complimentary simulation algorithm to Wang-Landau, in that, if detailed
%balance is ensured, the system is guaranteed to converge.  Unfortunately,
%the algorithm can be considerably more difficult to implement when
%compared to Wang-Landau: in particular, due to the infinite temperature
%transition matrix~\cite{wang2002transition}.  Also, the time necessary
%for the density of states to converge could be considerable
%due to the algorithm spending too much time exploring low energy states.
%Unlike WL, TMMC does not require a prior knowledge of the energy range of
%the system.
%
%Shell et al.~\cite{shell2003improved, shell2004flat} originially
%implemented Wang-Landau Transition Matrix Monte Carlo (WL-TMMC) in an
%effort to quickly explore the energies of a given system using WL then
%switch over to TMMC to guarantee convergence. Considerable effort has
%been spent trying to determine the flatness criteria and cutoff for the
%WL portion, but there remains no rigorous systematic way to determine
%these parameters~\cite{rane2013monte}.  The algorithm is first run for
%several trial simulations and/or `experience' is used to determine the
%best parameters for a users particular system of
%study~\cite{siderius2013use}.
\jpsays{I should look at some more 1/t-WL citations. I know a few more
that probably should be included here...}
Around the same time, the convergence rate for WL was being
explored~\cite{zhou2005understanding,lee2006convergence,
belardinelli2007wang}. It was found that utilizing modification factors
that decrease faster than $1/t^2$ leads to
nonconvergence~\cite{belardinelli2007fast}.  This leads to the so-called
$1/t$-WL algorithm which ensures that excess CPU time is not wasted by
continuing to perform calculations once the error in the density of
states becomes saturated~\cite{belardinelli2008analysis}. An effort to
reduce the number of user defined parameters of WL and formulate a
theory for why the method converges despite detailed balance being
violated although infrequently was also being undertaken.

@@

@@

%This work proposes a means for dynamically setting the update factor
%$\gamma_t$ that should robustly converge on a variety of systems.

>>>>

Wang and Landau introduced an algorithm that uses an update factor and
a statistical histogram to compute the density of states of a given
system~\cite{wang2001determining, wang2001efficient}.  While the method
is incredibly powerful, it has several user-defined
parameters~\cite{landau2004new}.  This can make its application to a
variety of systems something of an art-form since the user needs to
determine the ideal parameters for the particular system being studied.
Also, detailed balance is violated (although only for short periods of
time), ensuring that convergence is not guaranteed.  This raises
several questions such as when the method does converge, what rate will
the method converge to the correct density of states. Also, how can the
user appropriately decide what parameters to choose so that the
algorithm solves the problem in the most ideal way.

%Transition Matrix Monte Carlo~\cite{wang1999transition,
%swendsen1999transition, fitzgerald2000monte} became an attractive
%complimentary simulation algorithm to Wang-Landau, in that, if detailed
%balance is ensured, the system is guaranteed to converge.  Unfortunately,
%the algorithm can be considerably more difficult to implement when
%compared to Wang-Landau: in particular, due to the infinite temperature
%transition matrix~\cite{wang2002transition}.  Also, the time necessary
%for the density of states to converge could be considerable
%due to the algorithm spending too much time exploring low energy states.
%Unlike WL, TMMC does not require a prior knowledge of the energy range of
%the system.
%
%Shell et al.~\cite{shell2003improved, shell2004flat} originially
%implemented Wang-Landau Transition Matrix Monte Carlo (WL-TMMC) in an
%effort to quickly explore the energies of a given system using WL then
%switch over to TMMC to guarantee convergence. Considerable effort has
%been spent trying to determine the flatness criteria and cutoff for the
%WL portion, but there remains no rigorous systematic way to determine
%these parameters~\cite{rane2013monte}.  The algorithm is first run for
%several trial simulations and/or `experience' is used to determine the
%best parameters for a users particular system of
%study~\cite{siderius2013use}.
Around the same time, the convergence rate for WL was being
explored~\cite{zhou2005understanding,lee2006convergence,
belardinelli2007wang}. It was found that utilizing modification factors
that decrease faster than $1/t^2$ leads to
nonconvergence~\cite{belardinelli2007fast}.  This leads to the so-called
$1/t$-WL algorithm which ensures that excess CPU time is not wasted by
continuing to perform calculations once the error in the density of
states becomes saturated~\cite{belardinelli2008analysis}. An effort to
reduce the number of user defined parameters of WL and formulate a
theory for why the method converges despite detailed balance being
violated although infrequently was also being undertaken.

@@

\subsection{1/t Wang-Landau}
The Wang-Landau algorithm's error saturation can be corrected by
modifying the update factor such that it does not decrease too quickly.
Belardinelli and Pereyra raised the question as to whether the update
factor should be decrease by $\nicefrac12$ or some other factor for
each update~\cite{belardinelli2007fast}. They implemented a schedule
that enforced if the update factor $\gamma$ is less than
$\nicefrac{1}{t}$ then the update factor is modified by
$\nicefrac{1}{t}$ and the histogram is no longer tracked. Employing
this scheduler, they found that the error saturation is avoided since
the correct density of states is approached asymptotically as
$\nicefrac{1}{t}$~\cite{belardinelli2008analysis}. Zhou et. al further
confirmed that the WL algorithm never converges exponentially and
succesfully bounded the statistical error between
$\nicefrac{1}{t^{1/2}}$ and $\nicefrac{1}{t}$~\cite{zhou2008optimal}.

Schneider et. al outlines minor refinements to the 1/t WL algorithm
including $\nicefrac{N_E}{t}$ scaling and switching from standard WL to
1/t WL when the update factor $\gamma <
\nicefrac{N_E}{t}$~\cite{schneider2017convergence}. As per the original
1/t implementation~\cite{belardinelli2007fast}, all energy states $N_E$
are required to be visited at least once $H(E) \neq 0$ effectively
avoiding the concept of `flatness'.

@@
